manuscript_v1

### 2023-09-09 manuscript.Rmd
started rewriting

Animals assess a wide variety of environmental features at multiple scales, and make a series of hierarchical choices resulting in a preference for certain habitats over others, which is termed habitat selection *(Johnson, 1980; Cody, 1985)*. Broad-scale selection is reflected in species' geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter (Fretwell & Lucas, 1969; Johnson, 1980; Saab, 1999). This raises questions about the ways in which these preferences vary across species, space, time, etc. In birds, however, much of our knowledge on habitat selection is dominated by the breeding season---encompassing both the actual reproductive period, as well as the period leading up to mating (e.g., Danchin et al., 1998; Bosco et al., 2021). Moreover, most such studies target migratory species, whether it be in their breeding grounds (e.g., Danchin et al., 1998; Kleist et al., 2017), wintering grounds (usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017) or stopover sites (e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Cohen et al., 2021). This disproportionate attention is not unjustified: fitness outcomes are influenced via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and fitness has traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales.

Nevertheless, there is slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements and pressures during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014). In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in the post-breeding period. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a).

Among temperate forests, oak forests are widely believed to be the most biodiversity-rich, supporting a highly diverse arthropod community which in turn supports a larger vertebrate community. Although many migratory birds are known to become facultatively frugivorous in autumn (Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006), making them less reliant on the declining arthropods, European temperate forests have a low diversity of fruiting plants. Thus, the dietary options for post-breeding Central European forest birds are rather limited.

Under these various ecological pressures, competitive interactions may arise. In the presence of interspecific competition, species within the same guild would have smaller overall habitat breadths and overlaps, due to similar resource use patterns (Svärdson, 1949). On the other hand, in closed habitats such as woodlands, predators attack at a close range and predation is thus more important than feeding competition which is generally lower between species than within (Terborgh, 1990?). The absence of competition would allow species in the same guild to overlap more and take advantage of a shared feeding strategy. Hence, birds can be expected to select for habitats that maximise foraging efficiency and, eventually, survival. One such example is that of mixed-species foraging flocks, which not only allow higher and more efficient foraging rates, but also minimise costs of vigilance against predators (Sridhar et al., 2009).

Habitat selection patterns and strategies are therefore likely to be driven more strongly by characteristics of the habitat, than by dietary preferences. Bird abundances or densities are compared with habitat features such as canopy cover, heterogeneity, tree density, and species richness to analyse habitat selection (Vitz and Rodewald, 2007). Berg (1997) found that breeding bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*. Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. However, most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity (Fanta, 2007). While intense management and disturbance regimes do bring in heterogeneity, they do not facilitate biodiversity for obvious reasons. As a result, birds during the post-breeding period could show a strong foraging preference for oak forest patches, which not only support higher biodiversity but also have more heterogeneous canopies and understoreys than the structurally uniform coniferous forests.

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially used patch of Central European temperate mixed forest (*Branisovsky les*) close to České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure Norway spruce *Picea abies* plantations to mixed stands dominated by oaks (*Quercus* spp.). As such, the tree community consists mostly of conifers like Norway spruce, Scots pine *Pinus sylvestris*, European larch *Larix decidua* and silver fir *Abies alba*, and broad-leaved species such as pedunculate/European oak *Quercus robur*, sessile/Cornish oak *Q. petraea*, northern red oak *Q. rubra*, silver birch *Betula pendula*, quaking/Eurasian aspen *Populus tremula*, common/European beech *Fagus sylvatica*, rowan or European mountain ash *Sorbus aucuparia*, small-leaved lime or littleleaf linden *Tilia cordata*, common/European black alder *Alnus glutinosa*, Norway maple *Acer platanoides* and sycamore maple *A. pseudoplatanus*. The commonly found shrub species are alder buckthorn *Frangula alnus*, rowan *Sorbus aucuparia* and black/European elderberry *Sambucus nigra*, while ground layer vegetation is restricted to graminoids like sedges *Carex* spp. (mostly *C. brizoides*), bramble *Rubus* sp. and European blueberry *Vaccinium myrtillus*.

@@

Vegetation features potentially relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was recorded according to whether asphalt roads ('Road') and forest-to-farmland transitions ('Edge') occurred within 150 m from the centre of the point or not ('Interior'). This was done using an aerial map (Mapy.cz). At each of the 32 points, all tree and shrub species were identified and counted, and the vertical strata to which they belonged noted ('Canopy' or 'Understorey'). The species identities were used to calculate species richness and diversity (Shannon's diversity index; Shannon, 1948), while counts were used to calculate density per 100 m\^2. For other plants that classified as ground vegetation, only broad identifications were made (e.g., graminoids or mosses), and instead of counting them just the 'dominant' species were identified. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch (Berdugo et al., 2018; Cornelissen et al., 2007). Section 1 in Supplementary Material details the habitat characterisation methodology.

We also calculated the proportion of deciduous trees in the 'Canopy' as this has been shown to be a more important factor than measures of tree diversity (Berg, 1997). Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10--15 m of the centre, which were then analysed using ImageJ software (Rasband, 1997) by converting to binary black-and-white images and calculating the percentage of cover across the whole image (see Forester, 2011; Figure S1). The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These canopy measurements were originally taken once in each sampling week, but most deciduous trees bore leaves even in the last week of sampling. Hence, the 13 weekly measurements of both canopy variables for each sampling point were instead averaged to provide more accurate measures of both, and these were used in the analyses. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period (Section 2 in Supplementary Material).

@@

Point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate and separate two successive 'temporal units'. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point (Table 1).

@@

Each point count was associated with several variables such as Point ID, sampling day and week, observer, and other non-focal variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in) (see Table 2 in Supplementary Material). We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird's 'use' of the habitat. All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as 'spuhs'.

Prior to analysis, bird species were classified based on feeding guild as well as migration strategy. Feeding guild was determined usingGregory et al., 2007 (pt. Appendix 1) and Birds of the World (Billerman et al., 2020). Migration strategy was assigned following Czech ornithological data; species of which $\gt$ 75% of the Czech breeding population migrates either short or long distances (categories 3 and 4 in Cepák et al., 2008) were considered as complete migrants for the purposes of this study; those with 'regular and abundant wintertime occurrence in the Czech Republic' were considered partial migrants (category 2 in Cepák et al., 2008; see also Št'astnỳ and Hudec, 2011); and the residents of category 1 (Cepák et al., 2008) were considered true residents.

@@

To assess patterns of habitat selection by birds, we first excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. The measured habitat variables without the ones excluded due to correlations represented the pool of candidate predictors (Table 2 in Supplementary Material). This set of plausible predictors was then used to build generalised linear mixed models (GLMMs) following an information-theoretic approach (Burnham and Anderson, 1998). That is, the Akaike Information Criterion (AIC; Akaike, 1973, 1974), and more specifically the corrected AIC or AIC~C~ (Hurvich and Tsai, 1989), was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count, which was then used as the response variable. GLMMs were constructed using the *lme4* package (Bates et al., 2015), with the null model containing observer as a fixed effect (detections had 71:29 observer ratio) and sampling points and days as crossed random effects (model construction detailed later). Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function.

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficiently high number of overall detections, insectivores (rather, 'invertebrate-feeders') and omnivores. Models were built separately for these two guilds, using the *glmmTMB* package (Brooks et al., 2017) to avoid convergence issues. The data summarised at guild level was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables were added. Following this, the focal variables were finally added. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package (Bolker et al., 2020), with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser (Table 2); the guild-level final models were a Poisson GLMM (log link) for invertebrate-feeders and a negative binomial GLMM (log link) for omnivores.

For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Model diagnostics of all models were explored using the *DHARMa* package (Hartig, 2020) for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. All analyses were done using R version ***4.1.2 (R Core Team, 2021)***. Individual figures with multiple graphs were composed using the *patchwork* package for R (Pedersen, 2020).

@@

@@

@@

The common understorey species in our forest were *F. alnus* and *P. abies*, followed by *Q. robur/petraea* and *B. pendula*. Other species occurred rarely and even when they did, were sparse, isolated and in very few numbers. The presence of shrubs in the understorey is generally expected to be good for small passerines, but there was no detectable trend in bird abundance with understorey characteristics such as richness, diversity and density. One potential reason for this could be that the understorey characteristics we measured did not vary across a large enough range in our study site so as to be of significance to the birds. Hence, different results could be expected in other larger, less-degraded and naturally more diverse forests.

@@

@@

@@

>>>>

Animals assess a wide variety of environmental features at multiple scales, and make a series of hierarchical choices resulting in a preference for certain habitats over others, which is termed habitat selection *(Johnson, 1980; Cody, 1985)*. Broad-scale selection is reflected in species' geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter *(Fretwell & Lucas, 1969; Johnson, 1980; Saab, 1999)*. This raises questions about the ways in which these preferences vary across species, space, time, etc. In birds, however, much of our knowledge on habitat selection is dominated by the breeding season---encompassing both the actual reproductive period as well as the period leading up to mating *(e.g., Danchin et al., 1998; Bosco et al., 2021)*. Moreover, most such studies target migratory species, whether it be in their breeding grounds *(e.g., Danchin et al., 1998; Kleist et al., 2017)*, wintering grounds *(usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017)* or stopover sites *(e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Cohen et al., 2021)*. This disproportionate attention is not unjustified: reproductive success and offspring survival have traditionally been strongly linked with fitness, and fitness outcomes are influenced via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales *(e.g., Devries 2018, Wolfe 2014, Rodewald 2004, Winker 1995)*.

Nevertheless, there is slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements and pressures during the different seasons *(Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014)*. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in the post-breeding period. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas *(Uesugi and Murakami, 2007)*. Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators *(Anders et al., 1997)*. Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period *(Betts et al., 2008a)*.

Among temperate forests, oak forests are widely believed to be the most biodiversity-rich, supporting a highly diverse arthropod community which in turn supports a larger vertebrate community *(Kennedy and SOuthwood 1984)*. Although many migratory birds are known to become facultatively frugivorous in autumn *(Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006)*, making them less reliant on the declining arthropods, European temperate forests have a lower diversity of fleshy-fruit plants than North American ones *(Campbell 1982)*. Thus, the dietary options for post-breeding Central European forest birds are rather limited.

Under these various ecological pressures, competitive interactions may arise. In the presence of interspecific competition, species within the same guild would have smaller overall habitat breadths and overlaps, due to similar resource use patterns *(Svärdson, 1949)*. On the other hand, in closed habitats such as woodlands, predators attack at a close range and predation is thus more important than feeding competition which is generally lower between species than within *(Terborgh, 1990?)*. The absence of competition would allow species in the same guild to overlap more and take advantage of a shared feeding strategy. Hence, birds can be expected to select for habitats that maximise foraging efficiency and, eventually, survival. One such example is that of mixed-species foraging flocks, which not only allow higher and more efficient foraging rates, but also minimise costs of vigilance against predators *(Sridhar et al., 2009)*.

Habitat selection patterns and strategies are therefore likely to be driven more strongly by characteristics of the habitat, than by dietary preferences. Bird abundances or densities are compared with habitat features such as canopy cover, heterogeneity, tree density, and species richness to analyse habitat selection *(Vitz and Rodewald, 2007)*. *Berg (1997)* found that breeding bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*. Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. However, most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity *(Fanta, 2007)*. While intense management and disturbance regimes do bring in heterogeneity, they do not facilitate biodiversity for obvious reasons. As a result, birds during the post-breeding period could show a strong foraging preference for oak forest patches, which not only support higher biodiversity but also have more heterogeneous canopies and understoreys than the structurally uniform coniferous forests *(e.g., Bergner 2015)*.

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially used patch of Central European temperate mixed forest (*Branišovský les*) close to České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure Norway spruce *Picea abies* plantations to mixed stands dominated by oaks (*Quercus* spp.). For more details on the species composition of the vegetation community, see *Supplementary Material section*.

@@

Vegetation features potentially relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was recorded according to whether asphalt roads ('Road') and forest-to-farmland transitions ('Edge') occurred within 150 m from the centre of the point or not ('Interior'). This was done using an aerial map (Mapy.cz). At each of the 32 points, all tree and shrub species within the 30 m radius were identified and counted, and the vertical strata to which they belonged noted ('Canopy' or 'Understorey'). Species identities were used to calculate species richness and diversity *(Shannon's diversity index; Shannon, 1948)*, while counts were used to calculate density per 100 m^2^. For other plants that classified as ground vegetation, only broad identifications were made (e.g., graminoids or mosses), and instead of counting them only the 'dominant' species (covering roughly $\ge$ 50% ground area) were identified. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch *(Berdugo et al., 2018; Cornelissen et al., 2007)*. *Section 1 in Supplementary Material* details the habitat characterisation methodology.

We also calculated the proportion of deciduous trees in the 'Canopy' as this has been shown to be a more important factor than measures of tree diversity *(Berg, 1997)*. Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10--15 m of the centre, which were then analysed using ImageJ software *(Rasband, 1997)* by converting to binary black-and-white images and calculating the percentage of cover across the whole image *(see Forester, 2011; Figure S1)*. The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These canopy measurements were originally taken once in each sampling week, but most deciduous trees bore leaves even in the last week of sampling. Hence, the 13 weekly measurements of both canopy variables for each sampling point were instead averaged to provide more accurate measures of both, and these were used in the analyses. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period *(Section 2 in Supplementary Material)*.

@@

Point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate and separate two successive 'temporal units'. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point *(Table 1)*.

@@

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Each point count was associated with several variables *(see Table 2 in Supplementary Material)* such as Point ID, sampling day and week, observer, and other non-focal variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird's 'use' of the habitat.

All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as 'spuhs'. Prior to analysis, bird species were also classified based on feeding guild, which was determined using *Gregory et al., 2007 (pt. Appendix 1)* and *Birds of the World (Billerman et al., 2020)*.

@@

To assess patterns of habitat selection by birds, we first excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. The measured habitat variables, without those excluded due to correlations, represented the pool of candidate predictors *(Table 2 in Supplementary Material)*. This set of plausible predictors was then used to build generalised linear mixed models (GLMMs) following an information-theoretic approach *(Burnham and Anderson, 1998)*. That is, the Akaike Information Criterion *(AIC; Akaike, 1973, 1974)*, and more specifically the corrected AIC or AIC~C~ *(Hurvich and Tsai, 1989)*, was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count, which was then used as the response variable. GLMMs were constructed using the *lme4* package *(Bates et al., 2015)*, with the null model containing observer as a fixed effect (detections had *71:29* observer ratio) and sampling points and days as crossed random effects (model construction detailed later). Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function.

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficiently high number of overall detections, insectivores (rather, 'invertebrate-feeders') and omnivores. Models were built separately for these two guilds, using the *glmmTMB* package *(Brooks et al., 2017)* to avoid convergence issues. The data summarised at guild level was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables were added. Following this, the other, focal variables were finally added *(see Table 1)*. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package *(Bolker et al., 2020)*, with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser *(Table 2)*; the guild-level final models were a Poisson GLMM (log link) for invertebrate-feeders and a negative binomial GLMM (log link) for omnivores.

For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Model diagnostics of all models were explored using the *DHARMa* package *(Hartig, 2020)* for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. All analyses were done using R version ***4.1.1 (R Core Team, 2021)***. Individual figures with multiple graphs were composed using the *patchwork* package for R *(Pedersen, 2020)*.

@@

## Post-breeding birds track structural characteristics over species diversity

## Canopy heterogeneity as part of habitat heterogeneity

@@

## The surprising role of oak in post-breeding season

@@

## Understorey

The common understorey species in our forest were *F. alnus* and *P. abies*, followed by *Q. robur/petraea* and *B. pendula*. Other species occurred rarely and even when they did, were sparse, isolated and in very few numbers. The presence of shrubs in the understorey is generally expected to be good for small passerines *(Bergner 2015)*, but there was no detectable trend in bird abundance with understorey characteristics such as richness, diversity and density. One potential reason for this could be that the understorey characteristics we measured did not vary across a large enough range in our study site so as to be of significance to the birds. Hence, different results could be expected in other larger, less-degraded and naturally more diverse forests.

@@

## Changing with the seasons

@@

## Guilds

@@

## Conclusion
